Many-core applications to online track reconstruction in HEP experiments

### 2013-10-21 papers/2013/CHEP/paper.tex
english pass

\documentclass[letterpaper]{jpconf}
\usepackage{graphicx}
\usepackage{subfigure}
\usepackage{url}
\usepackage{xspace}
\usepackage{iopams}
\usepackage[pdftex,
      colorlinks=true,
      urlcolor=blue,       % \href{...}{...} external (URL)
      filecolor=green,     % \href{...} local file
      linkcolor=blue,       % \ref{...} and \pageref{...}
      pagebackref,
      pdfpagemode=UseNone,
      bookmarksopen=true]{hyperref}

@@

\begin{abstract}
  In this paper we describe updates to a previous study of latencies
  in GPU for potential trigger applications, where we measured the
  latency to transfer data to/from the GPU, exploring the timing of
  different I/O technologies on different GPU models.  Those studies,
  where a simplified track fitting algorithm was parallelized and run
  on a GPU, show that latencies of few tens of microseconds can be
  achieved to transfer and process packets of 4 kB of data, combining
  the modern Infiniband data transfer technology with direct access to
  GPU memory allowed by NVIDIA GPUDirect utilities.  We now have
  expanded our latency studies to include other multi-core systems
  (Intel Xeon Phi/MIC in addition to NVIDIA GPUs). We also discuss the
  implementation of a scaled-up version of the algorithm used at CDF
  for online track reconstruction - the SVT algorithm - as a realistic
  test-case for low-latency trigger systems using new computing
  architectures for LHC experiment.

\end{abstract}

\section{Introduction}
%%% Verbatim from NSS paper
Real-time event reconstruction plays a fundamental role in High Energy
Physics (HEP) experiments at hadron colliders.  Reducing the rate of
data to be saved on tape from millions to hundreds of events per
second is critical. To increase the purity of the collected samples,
rate reduction has to be coupled with an initial selection of the most
interesting events.  In a typical hadron collider experiment, the
event rate has to be reduced from tens of MHz to a few kHz.  The
selection system (trigger) is usually organized in multiple levels,
each capable of performing a finer selection on more complex physics
objects describing the event. Trigger systems usually comprise a first
level based on custom hardware, followed by one or two levels usually
based on farms of general purpose processors.  The possibility of also
using commercial devices at a low trigger level is very appealing:
they are subject to continuous performance improvements driven by the
consumer market, are less expensive than dedicated hardware, and are
easier to support.  Among the commercial devices, Graphic Processing
Units (GPUs) are of particular interest for online selections given
their impressive computing power (the latest NVIDIA~\cite{bib_nvidia}
GPU architecture, Kepler, exceeds Teraflop computing power); moreover,
high-level programming architectures based on C/C++ such as
\textsc{cuda}~\cite{bib_cuda} and \textsc{opencl}~\cite{bib_opencl}
make these devices more accessible to the general user.  The ultimate
goal of this study is to investigate the strengths and weaknesses of
GPUs when applied in a low latency environment, with particular
emphasis on the data transfer latency to/from the GPU and the
algorithm latency for processing on the GPU in a manner similar to a
typical HEP trigger application.

@@

We ported the track fitting part to parallel devices, as it is well suited to 
parallelization.

@@

\section{Experimental setup and data flow}
The many-core devices used in this study are listed in Table~\ref{tab_hwspecs}.

\begin{table}[!t]
  \centering
  \begin{tabular}{|l|c|c|c|c|}
    \hline
    Model & Tesla M2050 & Tesla K20m & GeForce GTX  Titan & MIC 5110P \\
    \hline
    \hline
    Performance (SP, GFlops) & 1030 & 3520 & 4500 & 2022 \\
    Memory bandwidth  (GB/s) & 148 & 208 & 288  & 320\\   
    Memory size (GB) & 3 & 5 & 6 & 8 \\
    Number of cores & 448 & 2496 & 2688 & 240 \\
    Clock speed (GHz) & 1.15 & 0.706 & 0.837 & 1.053 \\
    \hline
  \end{tabular}
  \caption{Capabilities of the many-core devices used in this study, according to the manufacturer's specifications.}
  \label{tab_hwspecs}
\end{table}

@@

\subsubsection{Breakdown of computing time}
In Fig.~\ref{fig:breakdown} we show the fractional time spent in
various parts of the algorithm for the embarrassingly parallel algorithm 
(on Intel MIC) and the parallel algorithm (on NVIDIA Titan GPU), as a 
function of the number of fits. On both accelerator cards the fractional times
are constant over 500 input events, where computing resources are saturated. 
Differently from MIC, the fit stage takes most of the time on the GPU: this 
could be explained with the intense memory access intrinsic in 
this part of the algorithm.  
\begin{figure*}[tbp]
\centering
\subfigure[]
{\label{fig:breakdown_MIC}
\includegraphics[width=0.45\linewidth]{figures/Mic5_perc_nk}}
\subfigure[]
{\label{fig:breakdown_GPU}
\includegraphics[width=0.45\linewidth]{figures/titan_perc}}
\caption{Breakdown of computing time for MIC (a) and the GTX Titan GPU
  (b). \fixme{we need a legend for these plots. Also can we make them
    the same size and aspect ratio.}}
\label{fig:breakdown}
\end{figure*}

>>>>

\documentclass[a4]{jpconf}
\usepackage{graphicx}
\usepackage{subfigure}
\usepackage{url}
\usepackage{xspace}
\usepackage{iopams}
\usepackage[pdftex,
      colorlinks=true,
      urlcolor=blue,       % \href{...}{...} external (URL)
      filecolor=green,     % \href{...} local file
      linkcolor=blue,       % \ref{...} and \pageref{...}
      pagebackref,
      pdfpagemode=UseNone,
      bookmarksopen=true]{hyperref}

@@

\begin{abstract}
  In this paper we describe updates to a previous study of latencies
  in GPU for potential trigger applications, where we measured the
  latency to transfer data to/from the GPU, exploring the timing of
  different I/O technologies on different GPU models.  Those studies,
  where a simplified track fitting algorithm was parallelized and run
  on a GPU, show that latencies of few tens of microseconds can be
  achieved to transfer and process packets of 4 kB of data, combining
  the modern Infiniband data transfer technology with direct access to
  GPU memory allowed by NVIDIA GPUDirect utilities.  We now have
  expanded our latency studies to include other multi-core systems
  (Intel Xeon Phi/MIC in addition to NVIDIA GPUs). We also discuss the
  implementation of a scaled-up version of the algorithm used at CDF
  for online track reconstruction - the SVT algorithm - as a realistic
  test-case for low-latency trigger systems using new computing
  architectures for LHC experiment and examine the
  complexity/performance trade-off in porting existing serial
  algorithms to many-core devices.
\end{abstract}

\section{Introduction}
%%% Verbatim from NSS paper
Real-time event reconstruction plays a fundamental role in High Energy
Physics (HEP) experiments at hadron colliders.  Reducing the rate of
data to be saved on tape from millions to hundreds of events per
second is critical. To increase the purity of the collected samples,
rate reduction has to be coupled with an initial selection of the most
interesting events.  In a typical hadron collider experiment, the
event rate has to be reduced from tens of MHz to a few kHz.  The
selection system (trigger) is usually organized in multiple levels,
each capable of performing a finer selection on more complex physics
objects describing the event. Trigger systems usually comprise a first
level based on custom hardware, followed by one or two levels usually
based on farms of general purpose processors.  The possibility of also
using commercial devices at a low trigger level is very appealing:
they are subject to continuous performance improvements driven by the
consumer market, are less expensive than dedicated hardware, and are
easier to support.  Among the commercial devices, Graphic Processing
Units (GPUs) are of particular interest for online selections given
their impressive computing power (the latest NVIDIA~\cite{bib_nvidia}
GPU architecture, Kepler, exceeds Teraflop computing power); moreover,
high-level programming architectures based on C/C++ such as
\textsc{cuda}~\cite{bib_cuda} and \textsc{opencl}~\cite{bib_opencl}
make these devices more accessible to the general user.  The ultimate
goal of this study is to investigate the strengths and weaknesses of
GPUs when applied in a low latency environment, with particular
emphasis on the data transfer latency to/from the GPU and the
algorithm latency for processing on the GPU in a manner similar to a
typical HEP trigger application, and to understand the cost/complexity
ratio of porting legacy serial code to many-core devices..

@@

We ported the track fitting as it is well suited to parallelization -
each track can be handled independently.

@@

\section{Experimental setup and data flow}
The many-core devices used in this study are listed in
Table~\ref{tab_hwspecs}. The GPUs include a less expensive gaming
class GPU (GTX) and ones optimized for scientific computing
(Tesla). The MIC corresponds to a Xeon Phi introduced in November 2012.

\begin{table}[!t]
  \centering
  \begin{tabular}{|l|c|c|c|c|}
    \hline
    Model & Tesla M2050 & Tesla K20m & GeForce GTX  Titan & MIC 5110P \\
    \hline
    \hline
    Performance (SP, GFlops) & 1030 & 3520 & 4500 & 2022 \\
    Memory bandwidth  (GB/s) & 148 & 208 & 288  & 320\\   
    Memory size (GB) & 3 & 5 & 6 & 8 \\
    Number of cores & 448 & 2496 & 2688 & 240 \\
    Clock speed (GHz) & 1.15 & 0.706 & 0.837 & 1.053 \\
    \hline
  \end{tabular}
  \caption{Capabilities of the many-core devices used in this study,
    according to the manufacturer's specifications. The first three
    are GPUs and the final is an Intel Xeon Phi.}
  \label{tab_hwspecs}
\end{table}

@@

\subsubsection{Breakdown of computing time}
In Fig.~\ref{fig:breakdown} we show the fractional time spent in
various parts of the algorithm for the embarrassingly parallel algorithm 
(on Intel MIC) and the parallel algorithm (on NVIDIA Titan GPU), as a 
function of the number of fits. On both accelerator cards the fractional times
are constant over 500 input events, where computing resources are saturated. 
Unlike the MIC, the fit stage takes most of the time on the GPU: this 
could be caused by the intense memory access frequency intrinsic to
this part of the algorithm.  
\begin{figure*}[tbp]
\centering
\subfigure[]
{\label{fig:breakdown_MIC}
\includegraphics[width=0.45\linewidth]{figures/Mic5_perc_nk}}
\subfigure[]
{\label{fig:breakdown_GPU}
\includegraphics[width=0.45\linewidth]{figures/titan_perc}}
\caption{Breakdown of computing time for MIC (a) and the GTX Titan GPU
  (b). Red corresponds to unpacking, green corresponds to generating
  hit combinations, blue is the linearized track fit, and magenta
  corresponds to offloading (MIC only)\fixme{we need a legend for
    these plots, and it doesn't work in B\&W. Also can we make them
    the same size and aspect ratio.} For MIC, combinations and fitting
  take the same amount of time for large number of events. For GPU,
  fitting dominates. }
\label{fig:breakdown}
\end{figure*}
